---
title: "Multivariate Adaptive Regression Splines (MARS)"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{mars}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(tidyverse)
```
# Multivariate Adaptive Regression Spline (MARS) Models

## Contents {#contents}

1 Getting started
2 The MARS algorithm
3 Preparing inputs
4 Methods

## 1 Getting started

### 1.1 Introduction
As discussed in Jerome H. Friedman's (1991) paper, the MARS (Multivariate Adaptive Regression Splines) is used to solve non-linear regression problems.This algorithm creates linear regression models using an aggregate of functions to make predictions using recursive partitioning. 


### 1.2 Setting up the package
Install the package from GitHub. First install and load the devtools package and then install mars with
```{r,eval=FALSE}
library(devtools)
install_github(repo = 'https://github.com/SFUStatgen/SFUStat360Projects',
               subdir = 'Projects/project-tauseef1/mars')
```

You can then load the package.
```{r,eval=FALSE}
library(mars)
devtools::load_all()
```

## 2 The MARS algorithm
The model we receive at the end from the `mars` algorithm is built using both `forward` and backward `stepwise` functions, it fits the data that we supply and creates a linear combination of the the basis functions.

### 2.1 Important Terminology
- `Multivariate` - Able to generate model based on several input variables (high
dimensionality).
- `Adaptive` - Generates flexible models in passes each time adjusting the model.
- `Regression` - Estimation of relationship among independent and dependent variables.
- `Spline` - A piecewise defined polynomial function that is smooth (possesses higher order
derivatives) where polynomial pieces connect.
- `Knot` - The point at which two polynomial pieces connect.

### 2.2 Applications of MARS
Spam filter is mentioned in the book as an example. High dimensions are handled well. Non-linear relationships modeled piecewise. Continuous dependent variable typically, but can be categorical

## 4 Preparing inputs 
Ozone is the response variable

```{r}
library(mars)
devtools::load_all()
library(stats); data(mtcars)
form <- formula(mpg~ wt+hp)
mc <- mars.control(Mmax=10)
mm <- mars(form, mtcars, mc)
```

## 5 Methods

### 5.1 Forward pass stepwise
At each step, `mars` adds the basis function which reduces the residual error the most. Always adds the basis function in `pairs`, both sides of knot. Calculate value for `knot` and function that fit the data, least squares. This is a greedy algorithm, the addition of model terms continues until the max number of terms in the model is reached.


### 5.2 Backward pass stepwise
Remove one term at a time from the model. Remove the term which increases the residual error the least. Continue removing terms until cross validation is satisfied. Use the `Generalized Cross Validation` (GCV) function for this purpose


### 5.3 Using  mars methods
This `mars` package has 4 methods which use the output from `mars()` function. The methods are are `plot()`, `predict.mars()`, `print()`, and `summary()`.

### 5.4 Predict
The `predictmars` takes new data and returns the predicted basis functions based on the fitted model.
```{r warning = FALSE, message = FALSE}
mc_predict <- mars.control(Mmax=2)
data <- mtcars[c("mpg", "wt", "hp")]
mtcars_mars <- mars(form, data[1:30,], mc_predict)
predict.mars(mtcars_mars)
```


### 5.5 Summary
The `summary.mars` prints a summary of the mars object with the function call. Also prints the summary of the hinges of each basis function.
```{r warning = FALSE, message = FALSE}
summary(mm)
```


### 5.6 Print
The `print.mars` method prints out a mars object
```{r warning = FALSE, message = FALSE}
print(mm)
```
### 5.7 Plot
Plots the fitted basis functions that depend on explanatory variable(main effects) or two explanatory variables (two-way interactions). 

Use `predict.lm()` to see the residual plots.
```{r}
plot
```

## Authors
Tauseef Kashtwari, Ibraheem Azad, Promit Chowdhury


## References
* Jerome H. Friedman. "Multivariate Adaptive Regression Splines."
Ann. Statist. 19 (1) 1 - 67, March, 1991. http://doi.org/10.1214/aos/1176347963
